---
title: "Log Transformation"
output: html_document
---

### Donald Kane
### October 25, 2021


```{r}
require(dplyr)
Hmisc::getHdata(nhgh)
d1 <- nhgh %>% 
  filter(sex == "female") %>% 
  filter(age >= 18) %>% 
  select(gh, ht) %>% 
  filter(1:n()<=1000)

gly <- d1$gh
ht <- d1$ht
```

# Glycohemoglobin

## Nomral Distribution

### Estimates of paramters

```{r}
# Method of Moments
xbar <- mean(gly)
xbar
s2 <- var(gly)
s2

```

### Overlay Estimated PDF onto Histogram

```{r}
# Method of Moments
x <- rep(0, 10, by = 0.1)
fglyg <- function(x){
  dnorm(x, xbar, s2)
}

hist(gly,
     freq = FALSE,
     main = "Normal Glycohemoglobin MM",
     breaks = 75)
curve(fglyg(x),
      add = TRUE,
      col = "red",
      lwd = 3,
      main = "Normal Glycohemoglobin MM")

# Maximum Likelihood
hist(gly, freq = FALSE, main = "")
curve(
    dnorm(x, coef(fit)[1], coef(fit)[2])
  , add = TRUE
  , col = "green"
  , lwd = 3
)
```

### Overlay Estimated CDF onto ECDF

```{r}
# Method of Moments


# Maximum Likelihood
plot(ecdf(gly), main = "")
curve(
    pnorm(x, coef(fit)[1], coef(fit)[2])
  , add = TRUE
  , col = "green"
  , lwd = 3
)
```

### QQ Plot

```{r}

```

### Estimated Median

```{r}

```

# Gamma Distribution

### Estimates of paramters

```{r}
# Method of Moments
xbar <- mean(gly)
s2 <- var(gly)

shape_hat <- xbar^2/s2
shape_hat
scale_hat <- s2/xbar
scale_hat

# Maximum Likelihood

```

### Overlay Estimated PDF onto Histogram

```{r}
# Method of Moments
fglyg <- function(x){
  dgamma(x,
         shape = shape_hat,
         scale = scale_hat)
}

x <- rep(0, 10, by = 0.1)
hist(gly,
     freq = FALSE,
     main = "Glycohemoglobin MM",
     breaks = 75)
curve(fglyg(x),
      add = TRUE,
      col = "red",
      lwd = 3,
      main = "Glycohemoglobin MM")


# Maximum Likelihood
hist(gly, freq = FALSE, main = "")
curve(
    dgamma(x, coef(fit)[1], coef(fit)[2])
  , add = TRUE
  , col = "green"
  , lwd = 3
)
```

### Overlay Estimated CDF onto ECDF

```{r}
# Method of Moments
Fglyg <- function(x){
  pgamma(x,
         shape = shape_hat,
         scale = scale_hat)
}

x <- rep(0, 10, by = 0.1)
plot(ecdf(gly),
     main = "Glycohemoglobin MM")
curve(Fglyg(x),
      add = TRUE,
      col = "red",
      lwd = 3,
      main = "Glycohemoglobin MM")

# Maximum Likelihood
plot(ecdf(bmi), main = "")
curve(
    pgamma(x, coef(fit)[1], coef(fit)[2])
  , add = TRUE
  , col = "green"
  , lwd = 3
)
```

### QQ Plot

```{r}

```

### Estimated Median

```{r}

```

# Weibull Distribution

### Estimates of paramters

```{r}

```

### Overlay Estimated PDF onto Histogram

```{r}

```

### Overlay Estimated CDF onto ECDF

```{r}

```

### QQ Plot

```{r}

```

### Estimated Median

```{r}

```

# Height




















