---
title: "Log Transformation"
output: html_document
---

### Donald Kane
### October 11, 2021

# Introduction

```{r message=FALSE}
library(ggplot2)
library(dplyr)
```

# Part 1

## Gamma Distribution

X ~ Gamma(shape = 3, scale = 1)

### Gamma Distribution Probability Density Function

```{r}
x <- seq(0,10, by = 0.1)
g_PDF <- dgamma(x, shape = 3, scale = 1)
g_CDF <- pgamma(x, shape = 3, scale = 1)
df <- data.frame(x, g_PDF, g_CDF)

ggplot(df, aes(x = x)) +
        geom_line(y = g_PDF) +
        geom_vline(aes(xintercept = mean(rgamma(1000, shape = 3, 
                                                scale = 1)),
                   color = "mean")) +
        geom_vline(aes(xintercept = qgamma(0.5, shape = 3, scale = 1),
                   color = "median")) +
        theme_classic() +
        ylim(0,0.5) +
        labs(
                title = "Gamma Distribution PDF",
                x = "x",
                y = "x ~ Gamma(shape = 3, scale = 1)"
        ) +
        scale_color_manual(name = "Statistics", 
                           values = c(median = "green", mean = "red"))
```

### Gamma Distribution Cummulative Density Function

```{r}
x <- seq(0,10, by = 0.1)
g_CDF <- pgamma(x, shape = 3, scale = 1)

ggplot(df, aes(x = x)) +
        geom_line(y = g_CDF) +
        geom_vline(aes(xintercept = mean(rgamma(1000, shape = 3, 
                                                scale = 1)),
                   color = "mean")) +
        geom_vline(aes(xintercept = qgamma(0.5, shape = 3, scale = 1),
                   color = "median")) +
        theme_classic() +
        ylim(0,1) +
        labs(
                title = "Gamma Distribution CDF",
                x = "x",
                y = "x ~ Gamma(shape = 3, scale = 1)"
        ) +
        scale_color_manual(name = "Statistics", 
                           values = c(median = "green", mean = "red"))
```

### Gamma Log Distribution Probability Density Function

```{r}
g_log <- log(rgamma(1000, shape = 3, scale = 1))

plot(density(g_log), type = "l", 
     main = "Log Gamma Probability Density Function")
abline(v = mean(g_log), col= 'red')
abline(v = median(g_log), col = "green")
legend('topleft', c('mean', 'median'),
       col = c('red', 'green'), lty = c(1, 1))
```

### Gamma Log Distribution Cummulative Density Function

```{r}
g_log <- log(rgamma(1000, shape = 3, scale = 1))

plot(ecdf(g_log), main = "Log Gamma Cummulative Density Function")
abline(v = mean(g_log), col= 'red')
abline(v = median(g_log), col = "green")
legend('bottomright', 
       c('mean', 'median'),
       col = c('red', 'green'), 
       lty = c(1, 1))

```

### Gamma Geometric and Arithmetic Means

```{r}
geometric = rep(NA, 1000)
arithmetic = rep(NA, 1000)
difference = rep(NA, 1000)

for (i in 1:1000){
  q = rgamma(100, shape = 3, scale = 1)
  
  arithmetic[i] = mean(q)
  geometric[i] = exp(mean(log(q)))
  difference[i] = abs(arithmetic[i] - geometric[i])
}

avgs = data.frame(arithmetic, geometric, difference)

ggplot(avgs, aes(x = geometric, y = arithmetic)) +
        geom_point() +
        geom_abline() +
        labs(x = 'Geometric Means',
             y = 'Arithmetic Means',
             title = 'Geometric vs. Arithmetic Means')

ggplot(avgs, aes(x = difference)) +
        geom_histogram() +
        labs(x = 'Difference',
             y = 'Frequency',
             title = 'Mean Differences')
```


## Log Normal Distribution

X ~ log Normal(mean = -1, standard deviation = 1)

### Log Normal Distribution Probability Density Function

```{r message=FALSE, warning=FALSE}
x1 <- seq(0,10, by = 0.1)
ln_PDF <- dlnorm(x1, -1, 1)
ln_CDF <- plnorm(x1, -1, 1)

df1 <- data.frame(x1, ln_PDF, ln_CDF)

ggplot(df1, aes(x = x1)) +
        geom_line(y = ln_PDF) +
        geom_vline(aes(xintercept = mean(rlnorm(1000, 
                                                meanlog = -1, sdlog = 1)),
                   color = "mean")) +
        geom_vline(aes(xintercept = qlnorm(0.5, meanlog = -1, sdlog = 1),
                   color = "median")) +
        theme_bw() +
        labs(
                title = "Log Normal Distribution PDF \nMean = -1, Standard Deviation = 1",
                x = "x",
                y = "x ~ Log Normal"
        ) +
        ylim(0,2)
        scale_color_manual(name = "Statistics", 
                           values = c(median = "green", mean = "red"))
```

### Log Normal Distribution Cummulative Density Function

```{r}
ggplot(df1, aes(x = x1)) +
        geom_line(y = ln_CDF) +
        geom_vline(aes(xintercept = mean(rlnorm(1000, 
                                                meanlog = -1, sdlog = 1)),
                   color = "mean")) +
        geom_vline(aes(xintercept = qlnorm(0.5, meanlog = -1, sdlog = 1),
                   color = "median")) +
        theme_bw() +
        labs(
                title = "Log Normal Distribution CDF \nMean = -1, Standard Deviation = 1",
                x = "x",
                y = "x ~ Log Normal"
        ) +
        ylim(0,1) +
        scale_color_manual(name = "Statistics", 
                           values = c(median = "green", mean = "red"))
```

### Log Normal Distribution Cummulative Density Function

```{r}
lg_norm <- log(rlnorm(1000, -1, 1))

plot(density(lg_norm), type = 'l', main = 'Log Normal Distribution Probability Density Function')
abline(v = mean(lg_norm), col = 'red')
abline(v = median(lg_norm), col = 'green')
legend('topleft', 
       c('mean', 'median'), 
       col = c('red', 'green'), 
       lty = c(1, 1))
```

### Log Normal Distribution Cummulative Density Function

```{r}
lg_norm <- log(rlnorm(1000, -1, 1))

plot(ecdf(lg_norm), main = "Log Normal Cummulative Density Function")
abline(v = mean(lg_norm), col= 'red')
abline(v = median(lg_norm), col = "green")
legend('topleft', 
       c('mean', 'median'),
       col = c('red', 'green'), 
       lty = c(1, 1))
```

### Log Normal Geometric and Arithmetic Means

```{r}
geometric = rep(NA, 1000)
arithmetic = rep(NA, 1000)
difference = rep(NA, 1000)

for (i in 1:1000){
  q = rlnorm(100, -1, 1)
  
  arithmetic[i] = mean(q)
  geometric[i] = exp(mean(log(q)))
  difference[i] = abs(arithmetic[i] - geometric[i])
}

avgs = data.frame(arithmetic, geometric, difference)

ggplot(avgs, aes(x = geometric, y = arithmetic)) +
        geom_point() +
        geom_abline() +
        labs(x = 'Geometric Means',
             y = 'Arithmetic Means',
             title = 'Geometric vs. Arithmetic Means')

ggplot(avgs, aes(x = difference)) +
        geom_histogram() +
        labs(x = 'Difference',
             y = 'Frequency',
             title = 'Mean Differences')
```


## Uniform Distribution

X ~ Uniform(min = 0, max = 12)

### Uniform Distribution Probability Density Function

```{r}
x2 <- seq(-1,13, by = 0.1)
u_PDF <- dunif(x2, min = 0, max = 12)
u_CDF <- punif(x2, min = 0, max = 12)
df2 <- data.frame(x2, u_PDF, u_CDF)

ggplot(df2, aes(x = x2)) +
        geom_line(y = u_PDF) +
        geom_vline(aes(xintercept = mean(runif(1000, 0, 12)),
                   color = "mean")) +
        geom_vline(aes(xintercept = qunif(0.5, 0, 12),
                   color = "median")) +
        theme_classic() +
        labs(
                title = "Uniform Distribution PDF",
                x = "x",
                y = "x ~ Uniform(0, 12)"
        ) +
        ylim(0,0.25) +
        scale_color_manual(name = "Statistics", 
                           values = c(median = "green", mean = "red"))
```

### Uniform Distribution Cummulative Density Function

```{r}
x2 <- seq(-1,13, by = 0.1)
u_CDF <- punif(x2, min = 0, max = 12)
df2 <- data.frame(x2, u_PDF, u_CDF)

ggplot(df2, aes(x = x2)) +
        geom_line(y = u_CDF) +
        geom_vline(aes(xintercept = mean(runif(1000, 0, 12)),
                   color = "mean")) +
        geom_vline(aes(xintercept = qunif(0.5, 0, 12),
                   color = "median")) +
        theme_classic() +
        labs(
                title = "Uniform Distribution CDF",
                x = "x",
                y = "x ~ Uniform(0,12)"
        ) +
        ylim(0,1) +
        scale_color_manual(name = "Statistics", 
                           values = c(median = "green", mean = "red"))

```

### Log Uniform Distribution Probability Density Function

```{r}
lg_unif <- log(runif(1000, 0, 12))

plot(density(lg_unif), type = 'l', main = 'Log Uniform Distribution Probability Density Function')
abline(v = mean(lg_unif), col = 'red')
abline(v = median(lg_unif), col = 'green')
legend('topleft', 
       c('mean', 'median'), 
       col = c('red', 'green'), 
       lty = c(1, 1))
```

### Log Uniform Distribution Cummulative Density Function

```{r}
lg_unif <- log(runif(1000, 0, 12))

plot(ecdf(lg_unif), main = "Log Normal Cummulative Density Function")
abline(v = mean(lg_unif), col= 'red')
abline(v = median(lg_unif), col = "green")
legend('topleft', 
       c('mean', 'median'),
       col = c('red', 'green'), 
       lty = c(1, 1))
```

### Log Normal Geometric and Arithmetic Means

```{r}
geometric = rep(NA, 1000)
arithmetic = rep(NA, 1000)
difference = rep(NA, 1000)

for (i in 1:1000){
  q = runif(100, 0, 12)
  
  arithmetic[i] = mean(q)
  geometric[i] = exp(mean(log(q)))
  difference[i] = abs(arithmetic[i] - geometric[i])
}

avgs = data.frame(arithmetic, geometric, difference)

ggplot(avgs, aes(x = geometric, y = arithmetic)) +
        geom_point() +
        geom_abline() +
        labs(x = 'Geometric Means',
             y = 'Arithmetic Means',
             title = 'Geometric vs. Arithmetic Means')

ggplot(avgs, aes(x = difference)) +
        geom_histogram() +
        labs(x = 'Difference',
             y = 'Frequency',
             title = 'Mean Differences')
```

# Part 3

What is the correct relationship between E[log(X)] and log(E[X])?

### Gamma

Based on the results below, we see the distribution is mostly in the negatives which allows us to say that the logarithm of the expected value of x is greater than the expected value of log(x) for the gamma distribution.

```{r}
expect_lg = rep(NA, 1000)
lg_expect = rep(NA, 1000)

for (i in 1:1000){
  q = rgamma(10000, shape = 3, scale = 1)
  expect_lg[i] = mean(log(q))
  lg_expect[i] = log(mean(q))
}

difference = expect_lg - lg_expect

hist(difference, freq = FALSE, breaks = 50,
     main = 'Difference between E[log(x)] and log(E[x])', 
     xlab = 'E[log(x)] - log(E[x])')
```

### Log Normal

Based on the results below, we see the distribution is mostly in the negatives which allows us to say that the logarithm of the expected value of x is greater than the expected value of log(x) for the log normal distribution.

```{r}
expect_lg = rep(NA, 1000)
lg_expect = rep(NA, 1000)

for (i in 1:1000){
  q = rlnorm(10000, -1, 1)
  expect_lg[i] = mean(log(q))
  lg_expect[i] = log(mean(q))
}

difference = expect_lg - lg_expect

hist(difference, freq = FALSE, breaks = 50,
     main = 'Difference between E[log(x)] and log(E[x])', 
     xlab = 'E[log(x)] - log(E[x])')
```

### Uniform

Based on the results below, we see the distribution is mostly in the negatives which allows us to say that the logarithm of the expected value of x is greater than the expected value of log(x) for the uniform distribution.

```{r}
expect_lg = rep(NA, 1000)
lg_expect = rep(NA, 1000)

for (i in 1:1000){
  q = runif(10000, 0, 12)
  expect_lg[i] = mean(log(q))
  lg_expect[i] = log(mean(q))
}

difference = expect_lg - lg_expect

hist(difference, freq = FALSE, breaks = 50,
     main = 'Difference between E[log(x)] and log(E[x])', 
     xlab = 'E[log(x)] - log(E[x])')
```
